---
title: "Marketing Campaign Data Analysis"
author: "Abdallah chidjou"
date: "`r format(Sys.Date(), '%B %d, %Y')`"
output: pdf_document
---


```{r}
# Abdallah Chidjou
# Citation: (source of help: Lecture note, googling in general, stackoverflow, and chatgpt)
```


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Introduction

In this analysis, we explore a dataset containing information from a marketing campaign. The dataset includes customer demographics, purchasing behavior, and responses to multiple campaigns. I aim to derive insights into customer characteristics and patterns that affect marketing success. 

- Describe the dataset and its variables.
- Perform statistical analyses and visualizations to understand key characteristics.
- Explore relationships between variables, focusing on purchasing behavior and campaign responses.

The dataset used includes the following variables:

- ID: Customer ID
- Year_Birth: Year of birth
- Education: Level of education
- Marital_Status: Marital status
- Income: Annual income
- Kidhome: Number of kids in the household
- Teenhome: Number of teenagers in the household
- Dt_Customer: Date when the customer became enrolled
- Recency: Days since last purchase
- MntWines, MntFruits, MntMeatProducts, MntFishProducts, MntSweetProducts, MntGoldProds: Money spent   on various product categories
- NumDealsPurchases, NumWebPurchases, NumCatalogPurchases, NumStorePurchases, NumWebVisitsMonth: Information about different purchasing channels and deals used
- AcceptedCmpX: Indicates whether the customer accepted a particular campaign (Cmp1, Cmp2, etc.)
- Complain: Whether the customer has complained in the past
- Z_CostContact: Fixed cost related to contact (likely constant)
- Z_Revenue: Revenue generated
- Response: Response to the most recent campaign
- (and others...; the folowing code illustrate the rest)


# Data Loading and Preparation

```{r}
library(dplyr)
library(ggplot2)
library(tidyr)
library(lubridate)
```



```{r}
# Load the dataset
marketing_data <- read.csv("marketing_campaign.csv", sep = "\t")

# Display the first few rows of the dataset
head(marketing_data)

# Data Cleaning

# Remove duplicates
marketing_data <- marketing_data %>% distinct()

# Visualize duplicates removal
cat("Number of rows after removing duplicates: ", nrow(marketing_data), "\n")

# Handling missing values - removing rows with NA values in the columns
marketing_data <- marketing_data %>% drop_na()


# Check for missing values in the dataset
missing_values <- is.na(marketing_data)

# Get a summary of missing values for each column
colSums(is.na(marketing_data))

# Convert date columns to appropriate format
marketing_data$Dt_Customer <- dmy(marketing_data$Dt_Customer)

# Check date conversion
cat("Data type of Dt_Customer after conversion: ", class(marketing_data$Dt_Customer), "\n")

# Remove unrealistic ages (e.g., customers older than 100 years)
marketing_data <- marketing_data %>% filter(2024 - Year_Birth <= 100)

# Visualize age filtering
ggplot(marketing_data, aes(x = 2024 - Year_Birth)) +
  geom_histogram(binwidth = 5, fill = "blue", color = "black") +
  theme_minimal() +
  labs(title = "Age Distribution After Removing Unrealistic Ages", x = "Age", y = "Count")

```



```{r}
# Convert the dataset from wide to long format so that we can easily boxplot all numerical columns
long_data <- marketing_data %>% 
  pivot_longer(cols = where(is.numeric), names_to = "Variable", values_to = "Value")

# Create boxplot for all numerical columns
ggplot(long_data, aes(x = Variable, y = Value)) +
  geom_boxplot() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  labs(title = "Boxplots of Numerical Columns", x = "Variables", y = "Values")

```



```{r}

# Handling outliers in Income by capping extreme values
income_quantiles <- quantile(marketing_data$Income, probs = c(0.01, 0.99))
marketing_data <- marketing_data %>% mutate(Income = ifelse(Income < income_quantiles[1], income_quantiles[1],
                                                            ifelse(Income > income_quantiles[2], income_quantiles[2], Income)))

# Visualize income outliers capping
ggplot(marketing_data, aes(x = Income)) +
  geom_histogram(binwidth = 5000, fill = "orange", color = "black") +
  theme_minimal() +
  labs(title = "Income Distribution After Handling Outliers", x = "Income", y = "Count")

```




```{r}
# Profit Calculation
# Assuming profit is derived from multiple categories of products
marketing_data <- marketing_data %>%
  mutate(Profit = MntWines * 0.3 + MntFruits * 0.2 + MntMeatProducts * 0.4 +
           MntFishProducts * 0.25 + MntSweetProducts * 0.15 + MntGoldProds * 0.35)

# Visualize profit distribution
ggplot(marketing_data, aes(x = Profit)) +
  geom_histogram(binwidth = 100, fill = "purple", color = "black") +
  theme_minimal() +
  labs(title = "Distribution of Customer Profit", x = "Profit", y = "Count")

```

# Data Exploration

## Descriptive Statistics

```{r descriptive-stats}
# Summary statistics for numeric variables
summary(marketing_data)
```


## Customer Demographics

We analyze customer demographics such as marital status and education level.

```{r marital-status}
# Bar plot for marital status
ggplot(marketing_data, aes(x = Marital_Status)) +
  geom_bar(fill = "green", color = "black") +
  theme_minimal() +
  labs(title = "Marital Status of Customers", x = "Marital Status", y = "Count")
```

# Data Engineering and Visualization

## Feature Engineering

- **Customer Tenure**: Calculate the number of days since the customer joined.

```{r feature-engineering}
marketing_data$Customer_Tenure <- as.numeric(difftime(Sys.Date(), marketing_data$Dt_Customer, units = "days"))

```

## Visualizations

- **Income Distribution**

```{r income-distribution}
ggplot(marketing_data, aes(x = Income)) +
  geom_histogram(binwidth = 5000, fill = "orange", color = "black") +
  theme_minimal() +
  labs(title = "Income Distribution of Customers", x = "Income", y = "Count")
```

- **Tenure Analysis**

```{r tenure-analysis}
ggplot(marketing_data, aes(x = Customer_Tenure)) +
  geom_histogram(binwidth = 100, fill = "cyan", color = "black") +
  theme_minimal() +
  labs(title = "Customer Tenure Distribution", x = "Days Since Joining", y = "Count")
```

# Data Analysis

## Spending Patterns

In this section, we investigate how customers spend across different product categories.

```{r spending-patterns}
# Spending on wine
ggplot(marketing_data, aes(x = MntWines)) +
  geom_histogram(binwidth = 50, fill = "purple", color = "black") +
  theme_minimal() +
  labs(title = "Distribution of Spending on Wine", x = "Amount Spent on Wine", y = "Count")
```

## Campaign Response Analysis

We now explore factors affecting responses to the most recent campaign.

```{r campaign-response}
# Response rates by income
ggplot(marketing_data, aes(x = Income, fill = factor(Response))) +
  geom_histogram(binwidth = 5000, position = "dodge", color = "black") +
  theme_minimal() +
  labs(title = "Campaign Response by Income", x = "Income", y = "Count", fill = "Response")
```

## Correlation Analysis: Income vs Profit

In this section, we explore the relationship between customer income and profit.

```{r income-profit-correlation}
# Scatter plot to visualize correlation between Income and Profit
ggplot(marketing_data, aes(x = Income, y = Profit)) +
  geom_point(alpha = 0.6, color = "blue") +
  theme_minimal() +
  labs(title = "Correlation between Income and Profit", x = "Income", y = "Profit")

# Calculate the correlation coefficient between Income and Profit
correlation <- cor(marketing_data$Income, marketing_data$Profit, use = "complete.obs")
cat("Correlation between Income and Profit: ", correlation, "\n")
```


## Income vs Profit Segmented by Marital Status

We further explore how the relationship between income and profit varies based on marital status.

```{r income-profit-marital-status}
# Scatter plot of Income vs Profit segmented by Marital Status
ggplot(marketing_data, aes(x = Income, y = Profit, color = Marital_Status)) +
  geom_point(alpha = 0.6) +
  theme_minimal() +
  labs(title = "Income vs Profit by Marital Status", x = "Income", y = "Profit", color = "Marital Status")

# Correlation coefficients by marital status, excluding groups with zero variance
correlation_by_status <- marketing_data %>%
  group_by(Marital_Status) %>%
  filter(sd(Income) > 0 & sd(Profit) > 0) %>%
  summarise(correlation = cor(Income, Profit, use = "complete.obs"))

print(correlation_by_status)
```

# Conclusion

In conclusion, the analysis shows several interesting insights into customer behavior and responses to marketing campaigns:

- The majority of customers are in the middle-age group, with notable spending patterns on wine and other products.
- There are clear differences in campaign responses based on income levels and family status.
- The correlation analysis reveals the relationship between customer income and profit, providing insights into how income influences profitability.
- The additional segmentation analysis indicates that the relationship between income and profit can vary significantly across different marital statuses, which could help in targeted marketing strategies.

