---
title: "California Housing Prices Data Analysis"
author: "Abdallah chidjou"
date: "`r format(Sys.Date(), '%B %d, %Y')`"
output: pdf_document
---


```{r}
# Abdallah Chidjou
# Citation: (source of help: Lecture note, googling in general, stackoverflow, and chatgpt)
```


```{r setup, include=FALSE}
# Install and load the reticulate package
if (!requireNamespace("reticulate", quietly = TRUE)) {
  install.packages("reticulate")
}
library(reticulate)
py_install("seaborn")
```


In this analysis, we explore a dataset containing information from California Housing Prices. The dataset includes the following features:

longitude: Geographic coordinate.
latitude: Geographic coordinate.
housing_median_age: Median age of the houses.
total_rooms: Total number of rooms.
total_bedrooms: Total number of bedrooms.
population: Population in the block.
households: Number of households.
median_income: Median income of residents.
median_house_value: Median house value.
ocean_proximity: Proximity to the ocean (categorical).

I aim to derive insights and patterns characterizing this dataset.

- Describe the dataset and its variables.
- Perform statistical analyses and visualizations to understand key characteristics.
- Explore relationships between variables.


```{python}
import pandas as pd
import matplotlib.pyplot as plt
import seaborn as sns
from sklearn.model_selection import train_test_split
from sklearn.linear_model import LinearRegression
from sklearn.metrics import r2_score, mean_squared_error

```



```{python}
# Load the dataset
file_path = 'California Housing Prices.csv'
housing_data = pd.read_csv(file_path)

# Inspect the dataset
initial_summary = {
    "Shape": housing_data.shape,
    "Columns": housing_data.columns.tolist(),
    "Data Types": housing_data.dtypes,
    "Missing Values": housing_data.isnull().sum()
}


for key, value in initial_summary.items():
    print(f"{key}:\n{value}\n")
```



```{python}

# Address Missing Values
# Fill missing values in the 'total_bedrooms' column with the median value
median_value = housing_data['total_bedrooms'].median()
housing_data['total_bedrooms'] = housing_data['total_bedrooms'].fillna(median_value)

# Generate Descriptive Statistics
numeric_summary = housing_data.describe()
numeric_summary

# Display updated missing values count to confirm resolution
updated_missing_values = housing_data.isnull().sum()

# Printing the updated missing values
updated_missing_values
```

```{python}
# Create boxplots for numeric columns
numeric_columns = housing_data.select_dtypes(include=['float64', 'int64']).columns
plt.title("Boxplots for Numeric Columns")

# Generate boxplots for each numeric column
plt.figure(figsize=(12, 8))
housing_data[numeric_columns].boxplot(rot=90)
plt.xlabel("Numeric Columns")
plt.ylabel("Values")
plt.tight_layout()
plt.show()

```
```{python}

```


```{python}
# Fix Outliers
# Use the IQR method to handle outliers
for column in numeric_columns:
    Q1 = housing_data[column].quantile(0.25)
    Q3 = housing_data[column].quantile(0.75)
    IQR = Q3 - Q1
    lower_bound = Q1 - 1.5 * IQR
    upper_bound = Q3 + 1.5 * IQR
    # Capping outliers
    housing_data[column] = housing_data[column].apply(lambda x: upper_bound if x > upper_bound else (lower_bound if x < lower_bound else x))

# Recheck boxplots after fixing outliers
plt.figure(figsize=(12, 8))
plt.title("Boxplots for Numeric Columns After Fixing Outliers")

# Generate updated boxplots for each numeric column
housing_data[numeric_columns].boxplot(rot=90)
plt.xlabel("Numeric Columns")
plt.ylabel("Values")
plt.tight_layout()
plt.show()

```









```{python}
# Scatter Plot: Median House Value vs. Median Income
plt.figure(figsize=(10, 6))
plt.scatter(housing_data['median_income'], housing_data['median_house_value'], alpha=0.5)
plt.title("Median House Value vs. Median Income")
plt.xlabel("Median Income")
plt.ylabel("Median House Value")
plt.grid(True)
plt.show()

```

```{python}
# Histogram: Distribution of Median House Value
plt.figure(figsize=(10, 6))
plt.hist(housing_data['median_house_value'], bins=30, edgecolor='k')
plt.title("Distribution of Median House Values")
plt.xlabel("Median House Value")
plt.ylabel("Frequency")
plt.show()
```

```{python}
plt.figure(figsize=(12, 8))
numeric_only_data = housing_data.select_dtypes(include=['float64', 'int64'])
correlation_matrix = numeric_only_data.corr()
sns.heatmap(correlation_matrix, annot=True, cmap="coolwarm", fmt=".2f")
plt.title("Correlation Heatmap of Features")
plt.show()
```



```{python}
# Boxplot: Median House Value by Ocean Proximity
plt.figure(figsize=(10, 6))
sns.boxplot(x='ocean_proximity', y='median_house_value', data=housing_data)
plt.title("Median House Value by Ocean Proximity")
plt.xlabel("Ocean Proximity")
plt.ylabel("Median House Value")
plt.xticks(rotation=45)
plt.show()
```




